---
title: "Environmental and geographic standardisation case study"
author: "G. Antell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: habitat-case-study-refs.bib
vignette: >
  %\VignetteIndexEntry{Environmental and geographic standardisation case study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

Palaeobiologists often track ecological parameters through time, where, as the package vignette illustrates, it is crucial to standardise differences in spatial coverage between time steps. However, it is equally important to apply spatial standardisation for comparisons between environment or habitat types, even if data occur within one time step.  For instance, consider two habitat types that cover different geographic extents. To determine fairly which habitat hosts more diverse communities, one must account for the difference in spatial coverage. By first subsampling each habitat type into equivalent spatial units, one could then directly and fairly compare richness or other ecological metrics between environments.

In the marine realm, one of the most important environmental axes of ecological differentiation is water depth. Diversity, geographic range size, and other aspects of community structure vary along a gradient of near-shore to deep-water environments. Here, we divide this environmental spectrum coarsely into shallow (proximal) vs. deep (distal) marine habitats. 'Shallow' habitats encompass deltaic, (sub)tidal, and shelf margin environments. 'Deep' habitats include continental slope and offshore settings, generally beginning at 140 m [@Duran2018].

The ecological attribute this case study focusses on is geographic range size, a trait with widespread importance. For instance, range size determines extinction risk to a large degree [@Harnik2012; @Saupe2015]. Range size might influence origination propensity as well, although researchers dispute whether this relationship is positive [@Rosenzweig1995; @Maurer2001], negative [@Hansen1980; @Jablonski2003], or unimodal [@Gaston2003; @Allmon2014]. 

The environmental niche breadth of a taxon strongly controls its geographic distribution; the maximum attainable extent of a taxon's range size is the extent of occurrence of that taxon's suitable habitat. Therefore, for environments within a given ecological niche breadth, a larger geographic distribution of those habitats permits taxa to occupy larger ranges. Taxa may be more restricted than suitable environmental conditions would predict (realised niche breadth), due to biotic or historic factors, but taxa cannot exceed the bounds set by environmental suitability. Since environmental characteristics are more homogeneous in the cold, dark waters of the deep sea than in the topologic complexity of shallow settings, ecologists have long theorised geographic ranges to be larger for deep-dwelling than shallow-water species [@Coe1946; @McClain2010; @Costello2017]. Most empirical tests of this hypothesis have relied upon modern data, however, with relatively little evidence from periods of Earth history when ocean circulation and continental configuration were entirely unlike those of today.

The following sections walk through an analysis to test the expectation geographic ranges are wider for deep- than shallow-water benthic taxa, using brachiopod occurrences from the Silurian (444-419 Ma) recorded in the [Paleobiology Database](https://paleobiodb.org/) (PBDB). Brachiopoda was a diverse clade that made up a large component of Paleozoic benthic communities [@Close2020; @Rojas2021]. In the Paleozoic, much of the world's landmass had coalesced at southermost latitudes into Gondwana, with the remaining continents scattered at the equator and separated by the Iapetus, Paleo-Tethys, Rheic, and Panthalassic oceans (map below). Hence, the geometry of a cosmopolitan marine species' range from the Silurian would differ distinctly from a cosmopolitan range today.

![Silurian palaeogeography, as reconstructed by the PALEOMAP PaleoAtlas project [@Scotese2016].](PALEOMAP_atlas_map75a_Llandovery_430.png)

This document consists of four parts, with `divvy` helper functions in `R` code chunks to analyse geographic range size disparities between marine environments.

1. Formatting occurrences (point data)
1. Developing environmental categories (raster data)
1. Spatially subsampling to regions of equal area and dispersion
1. Hypothesis testing using bootstrapped null simulations

# Prepare data

PBDB downloads of 1) all marine collections and 2) all Brachiopoda occurrences with accepted names at genus-ranking or below are available as attached datasets in the `divvy` package. Both datasets are subset to relevant columns and are cleaned to remove records missing coordinates. For more information, load the package and query the data objects from the console with `?collSilur` and `?occSilur`. The script to download and vet the raw data is available as `format-data.R` within the GitHub repository [data-raw folder (link)](https://github.com/GwenAntell/divvy/tree/main/data-raw).

```{r imports, message=FALSE}
# terra and sf have all the necessary helper functions for spatial operations
library(terra) 
library(sf)
library(divvy) 
data(collSilur)
data(occSilur)
```

The taxonomic hierarchy and stratigraphic ranges of the occurrence data are standardised against the Sepkoski Compendium already, following the `fossilbrush` package vignette [@fossilbrush]. A further element of cleaning is removal of taxa that occur only once ('space-time singletons' *sensu* @Antell2020). These records contain insufficient information to contribute to geographic range-size calculations.
```{r remove singletons}
nmFreq <- table(occSilur$genus)
sss <- nmFreq[nmFreq == 1] |> names()
sssRows <- occSilur$genus %in% sss
occSilur <- occSilur[ !sssRows, ]
```

To disregard trivial differences in coordinate values between points from the same or adjacent localities, the next data treatment is conversion of vector (point) data to raster grid cells. For each PBDB record, the ID and centroid coordinates of its encompassing raster cell are saved in new columns. These cell values will be used in spatial calculations later to improve computation time substantially compared to operations on vector point data.
```{r rasterise}
# work in Equal Earth projected coordinates
rWorld <- rast()
prj <- 'EPSG:8857'
rPrj <- project(rWorld, prj, res = 200000) # metre units, so ca. 2 degree resolution
values(rPrj) <- 1:ncell(rPrj)

xyCartes <- c('paleolng','paleolat')
xyCell   <- c('cellX','cellY')

# collection data
llColl <- vect(collSilur, geom = xyCartes, crs = 'epsg:4326')
prjColl <- project(llColl, prj)
collSilur$cell <- cells(rPrj, prjColl)[,'cell']
collSilur[, xyCell] <- xyFromCell(rPrj, collSilur$cell)

# taxon occurrences
llOccs <- vect(occSilur, geom = xyCartes, crs = 'epsg:4326')
prjOccs <- project(llOccs, prj)
occSilur$cell <- cells(rPrj, prjOccs)[,'cell']
occSilur[, xyCell] <- xyFromCell(rPrj, occSilur$cell)
```

# Classify substrate environment

To compare ecological parameters between deep and shallow habitat types, the 23 categories in the PBDB `environment` field must first be reclassified to these binary options. The following division of marine environments is copied from scripts of @Antell2020, who adapted the classification of @Nurnberg2013.
```{r define environments}
proxTerms <- c("coastal indet.", "delta front", "delta plain",
               "deltaic indet.", "estuary/bay", "foreshore", "interdistributary bay",
               "lagoonal", "lagoonal/restricted shallow subtidal",
               "marginal marine indet.", "open shallow subtidal", 
               "fluvial-deltaic indet.", "paralic indet.", 
               "peritidal", "prodelta", "sand shoal",
               "shallow subtidal indet.", "shoreface", 
               "transition zone/lower shoreface",
               "intrashelf/intraplatform reef", "reef, buildup or bioherm",
               "perireef or subreef", "platform/shelf-margin reef") 
deepTerms <- c("basinal (carbonate)", "basinal (siliceous)", 
               "basinal (siliciclastic)", "deep-water indet.", 
               "deep subtidal indet.", "deep subtidal ramp",
               "deep subtidal shelf", "offshore", "offshore indet.",
               "offshore shelf", "slope", "submarine fan", "offshore ramp", 
               "basin reef", "slope/ramp reef") 
```

The most accurate environmental information will be that associated directly with occurrence records. Where available, that is what this script will use. However, there are many instances of blank of indeterminate 'environment' field. In those cases, a way to infer probable environment is to determine the predominant environment in the local grid cell based on the environment recorded with any PBDB collections in that cell. A collection in the PBDB can contain any number of occurrences of any number of clades but is tied to a point coordinate - therefore, collections are a rich source of environmental data that is largely independent of occurrences for an individual clade. 

`divvy` provides the `classRast` function to convert environmental occurrence points to a raster grid with a dominant environmental class in each cell. The template raster passed to `classRast` here has the same resolution, extent, and coordinate reference system as above. Here, the function classifies grid cells as proximal or deep-water whenever more than 80% of collections are affiliated with a single environment.
```{r classify collections enviro, fig.width=5.5, fig.align='center'}
collSilur$bath <- NA
collSilur$bath[collSilur$environment %in% proxTerms] <- 'prox.' 
collSilur$bath[collSilur$environment %in% deepTerms] <- 'deep'
bathRast <- classRast(rPrj, dat = collSilur, cutoff = 0.8,
                      xy = xyCell, env = 'bath')
plot(bathRast, col = c("#1B9E77", "#D95F02", "black"))
```
A plot of the environmental distribution inferred from collection data traces the mid-latitude continents, albeit with gaps. Very few occurrences appear along the coasts of Gondwana, perhaps largely due to the under-sampling and under-reporting of the corresponding modern continents.

```{r classify occs enviro}
# create a new 'bathCell' column to note the collections-inferred environment
occSilur$bathCell <- occSilur$bath <- extract(bathRast, prjOccs)[['mainClass']]

# refine the enviro where possible based on observations directly linked to occs
occSilur$bath[occSilur$environment %in% proxTerms] <- 'prox.' 
occSilur$bath[occSilur$environment %in% deepTerms] <- 'deep'
```

A quick inspection reveals roughly equal numbers of occurrences in each environment. However, there are twice as many shallow as deep-water grid cells as indicated by all marine collections.
```{r relative enviro abundance}
# inspect: what proportion of environments associated with individual records
# match environmental inferrences from nearby collections' data?
which(occSilur$bath == occSilur$bathCell) |> 
  length()/nrow(occSilur)
table(occSilur$bathCell)
table(occSilur$bath)
# the majority of inferences do match, and most of the discrepancies are where
# more precise environmental information was available directly from occurrences

freq(bathRast, bylayer = FALSE)
```

# Spatial standardisation

Regionally subsample each occurrence subset. 
```{r subsample, error=TRUE}
deep <- occSilur[occSilur$bath == 'deep', ] |>
  uniqify(taxVar = 'genus', xy = xyCell)
prox <- occSilur[occSilur$bath == 'prox.', ]|>
  uniqify(taxVar = 'genus', xy = xyCell)
# the divvy uniqify() function removes duplicate taxon-cell combinations
# for smaller memory use and faster calculations below

reps <- 500
siteQuota <- 15
siteCol <- 'cell'
r <- 1000 # radial distance in km

set.seed(8)
sampD <- cookies(dat = deep,
                 xy = xyCell, iter = reps,
                 nSite = siteQuota,
                 siteId = siteCol, r = r,
                 weight = TRUE,
                 crs = prj, output = 'full')

set.seed(9)
sampP <- cookies(dat = prox,
                 xy = xyCell, iter = reps,
                 nSite = siteQuota,
                 siteId = siteCol, r = r,
                 weight = TRUE,
                 crs = prj, output = 'full')
```

# Analysis across environments

Determine what the results would conclude if analysis were based on global data without subsampling.
```{r non-subsampled results}
unsampD <- sdsumry(deep, 'genus', xyCell, prj, quotaQ = 0.95, quotaN = 1000, omitDom = TRUE)
unsampP <- sdsumry(prox, 'genus', xyCell, prj, quotaQ = 0.95, quotaN = 1000, omitDom = TRUE)
round(unsampD, 2)
round(unsampP, 2)
# paste0('deep ', round(unsampD[,'SCOR'], 2), '; proximal ', round(unsampP[,'SCOR'],2))
# [1] "deep 16.89; proximal 11.06"
# diversity roughly equivalent (exactly the same with rarefaction, but slightly higher 
# values in deep: 214 vs 201 total taxa, 192 to 184 taxa point-estimated from SQS)
```

Calculate summary variables for each subsample. Check how well various metrics correlate with each other across subsample replicates, e.g. total number of occurrences or taxa, or MST of subsample sites.
```{r var corrs}
# warning - very slow!! option to save and read in these data objects instead
pt1 <- proc.time()
metaD <- sdsumry(sampD, 'genus', xyCell, prj, quotaQ = 0.6, quotaN = 100, omitDom = TRUE)
metaP <- sdsumry(sampP, 'genus', xyCell, prj, quotaQ = 0.6, quotaN = 100, omitDom = TRUE)
pt2 <- proc.time()
(pt2-pt1)/60
#       user     system    elapsed 
# 26.7666667  0.2131667 27.2441667 
save(metaD, metaP, file = 'DP-subsample-metadata.RData')
load('DP-subsample-metadata.RData')

# need to check for infinity values of SCOR (when a taxon present in all cells)
# removing subsamples with these cases means sample-sizes may differ
# so remove an equal number of cases from deep/prox data sets
infScor <- is.infinite(metaD$SCOR)
nInf <- sum(infScor)
metaD <- metaD[ !infScor, ]
# pick the same number of rows to remove from the prox dataset
set.seed(10)
rows2toss <- sample(1:reps, nInf)
metaP <- metaP[-rows2toss, ]

# metaD$occAvg <- metaD$nOcc / metaD$nTax
# metaP$occAvg <- metaP$nOcc / metaP$nTax
  
# check out correlations between all pairs of variables
vars <- c('nTax','nOcc','minSpanTree') 
# vars <- c('nTax', 'SQSdiv', 'CRdiv','evenness')
cor(metaD[,vars], method = 'kendall')
cor(metaP[,vars], method = 'kendall')
```

```{r test diffs by enviro}
# NB: sorting strongly condenses the range of test statistic values
# and shifts from 0.5 to slightly less (because introduces ties (?))
nullSim <- function(dat1, dat2, vTest, vSort = NULL){
  # shuffle labels between test groups
  allDat <- rbind(dat1, dat2)
  n <- nrow(allDat) / 2
  hlf <- sample(sample(1:n*2), n, replace = FALSE)
  dat1null <- allDat[ hlf,]
  dat2null <- allDat[-hlf,]
  
  # optionally sort by second (sampling) variable
  if (! is.null(vSort)){
    ordr1 <- order(dat1null[, vSort])
    ordr2 <- order(dat2null[, vSort])
    dat1null <- dat1null[ordr1,]
    dat2null <- dat2null[ordr2,]
  }
  
  # (re)calculate test statistic
  v1null <- dat1null[, vTest]
  v2null <- dat2null[, vTest]
  sum(v2null > v1null) / n
}
# direction of test hypothesis here assumes group 2 values > group 1
# simulation also assumes groups 1 and 2 have equal sample size (n subsamples)

obsVsNull <- function(dat1, dat2, vTest, vSort = NULL, nullN){
  if ( !is.null(vSort) ){
    ordr1 <- order(dat1[, vSort])
    ordr2 <- order(dat2[, vSort])
    dat1 <- dat1[ordr1,]
    dat2 <- dat2[ordr2,]
  }
  h1obs <- dat2[, vTest] > dat1[, vTest]
  obsTestStat <- sum(h1obs) / nrow(dat1)
  nullDist <- replicate(nullN, nullSim(dat1, dat2, vTest, vSort))
  
  # hist(nullDist, xlim = c(0.4, 0.7)) # plot
  # abline(v = obsTestStat, col = 'blue')
  ci95 <- c(0.025, 0.975)
  null95 <- quantile(nullDist, ci95)
  names(null95) <- c('nullLwrCI', 'nullUprCI')
  # at what quantile of the null dist does the empirical test statistic lie?
  testQuantl <- sum(obsTestStat > nullDist) / nullN
  c(null95, 'obsTestStat' = obsTestStat, 'obsQuantl' = testQuantl)
}

# are shallows more diverse? (H: prox > deep) - only if estimated with SQS
obsVsNull(metaD, metaP, vTest = 'SCOR', nullN = 1000) 
obsVsNull(metaD, metaP, vTest = 'nTax', nullN = 1000) 
obsVsNull(metaD, metaP, vTest = 'SQSdiv', nullN = 1000) 
obsVsNull(metaD, metaP, vTest = 'CRdiv', nullN = 1000)
```

# References
